Synthesizing Quotient Lenses

### 2018-03-15 papers/QuotientLenses/main.tex
change color to dkblue and respond to B's commments

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkred}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,collapse,perm,squash,normalize,using,ins,del,lens,let,get,put,rquot,lquot},
emphstyle=\upshape\sffamily\color{dkred}, 
columns=fullflexible,
escapechar=\#,
mathescape, 
xleftmargin=1.5em,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
}
\makeatletter
     \let\lst@oldvisiblespace\lst@visiblespace
     \def\lst@visiblespace{\,\lst@oldvisiblespace\,}
\makeatother

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\textcolor{dkred}{\ensuremath{\mathsf{#1}}}}
\newcommand{\collapse}[2]{\ensuremath{\kw{collapse} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}(#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}
\newcommand{\bsep}{\ \ \sep{} \ \ }

@@

This paper applies this idea to the problem of
synthesizing {\em quotient lenses}~\cite{quotientlenses}.
Quotient lenses are lenses in which the lens laws are
loosened so that they hold modulo an equivalence relation on the source and
target data respectively; in this paper we are concerned with {\em bijective
quotient lenses} which are lenses for which Equation \ref{bijectivelenslaws}
holds modulo equivalence relations $\equiv_S$ and $\equiv_T$ defined on the source
and target data respectively:\bcp{Not too convinced about the colors.  The
  fact that lens expressions mix black and red is distracting -- my eyes
  can't easily tell which bits of a larger expression are lenses.  I would
  suggest either:
  \begin{itemize}
  \item get and put italicized, plus
  \item either and lens expressions (completely) in some subtle color (maybe
  brown), or simply black.
  \end{itemize}
}
\begin{equation}\label{quotientlenslaws}
\ell.\get \; (\ell.\lput \; t) \equiv_T t \text{, and } \ell.\lput \; (\ell.\get
\; s) \equiv_S s
\end{equation}
Quotient lenses are useful in situations where a programmer wishes for the
transformation defined by a lens to have the same behavior on data that differ
only in inessential details. For instance, a programmer may wish to ``quotient
away'' the number of white space characters between data items, the
capitalization of various strings, the sequence of fields in a record, or the
order of items in a list.

@@

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the edges,''
just like Boomerang's quotient lenses. 
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{QREs}
\caption{QRE lens from source $P$ to target $Q$.  $P$ and $Q$ are QREs, each
  consisting of a ``whole'' set ($W(P)$ and $W(Q)$) and a ``kernel'' part
  ($K(P)$ and $K(Q)$) Between the two kernels is a plain bijective lens.
  The \kw{get} function of the whole lens takes an argument from $W(P)$,
  applies the canonizer for $P$ to obtain a canonical representative in
  $K(P)$, then applies the \kw{get} of the plain lens, yielding an element
  of $K(Q)$, which is a subset of $W(Q)$.  The \kw{put} function does the
  reverse, mapping from $W(Q)$ to $K(P)$ (hence $W(P)$.  This lens is in
  normal form, with canonizers (specified by QREs) at the outer edges and an
  ordinary bijective lens in the center. \BCP{Both the figure and the
    caption are new---what do you think??}}\saz{I think that this figure is a
  big improvement.  One tiny possible tweak: color the \kw{get} and \kw{put}
  keywords in the same dark red as in the text.}\bcp{Can do, if we keep the
  color in the text :-).  How are we going to decide about that?}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: Q_1 \Leftrightarrow Q_2$ where $Q_1$ and $Q_2$
are QREs. In the {\em get} direction, a QRE lens $q_1: Q_1 \Leftrightarrow Q_2$ uses the
source QRE $Q_1$ to compute a canonical representative for the data modulo the
equivalence relation defined by $Q_1$ and then applies the \kw{get} function of a
bijective string lens $\ell_1$ to this representative. In the {\em put} direction, $q_1$
operates similarly, but using the QRE $Q_2$ and the $\lput$ function of
$\ell_1$.

@@

\BCP{Wrote the rest---comments?}
Looking ahead, we would like to investigate whether the techniques proposed
here can be applied to synthesizing quotiented variants of richer classes of
lenses, such as ``classic'' asymmetric lenses~\cite{FosterLenses} or
symmetric lenses~\cite{symmetriclenses}.  The main technical issues are
developing synthesis procedures for the unquotientied variants of a given
class---we are currently attacking the case of symmetric lenses---and
showing that quotiented lenses in this class can be normalized along the
lines of Theorem~\ref{normal form}.

>>>>

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,collapse,perm,squash,normalize,using,ins,del,lens,let,get,put,rquot,lquot},
emphstyle=\upshape\sffamily\color{dkblue}, 
columns=fullflexible,
escapechar=\#,
mathescape, 
xleftmargin=1.5em,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
}
\makeatletter
     \let\lst@oldvisiblespace\lst@visiblespace
     \def\lst@visiblespace{\,\lst@oldvisiblespace\,}
\makeatother

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\textcolor{dkblue}{\ensuremath{\mathsf{#1}}}}
\newcommand{\collapse}[2]{\ensuremath{\kw{collapse} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}(#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}
\newcommand{\bsep}{\ \ \sep{} \ \ }

@@

This paper applies this idea to the problem of
synthesizing {\em quotient lenses}~\cite{quotientlenses}.
Quotient lenses are lenses in which the lens laws are
loosened so that they hold modulo an equivalence relation on the source and
target data respectively; in this paper we are concerned with {\em bijective
quotient lenses} which are lenses for which Equation \ref{bijectivelenslaws}
holds modulo equivalence relations $\equiv_S$ and $\equiv_T$ defined on the source
and target data respectively:\bcp{Not too convinced about the colors.  The
  fact that lens expressions mix black and red is distracting -- my eyes
  can't easily tell which bits of a larger expression are lenses.  I would
  suggest either:
  \begin{itemize}
  \item get and put italicized, plus
  \item either and lens expressions (completely) in some subtle color (maybe
  brown), or simply black.
  \end{itemize}
}\saz{I agree that the coloring can be a bit distracting, but I do think that
  highlighting the keywords in the code helps for parsing--maybe using dark blue
  is better? Why do ``get'' and ``put'' deserve special treatment as keywords,
  compared to, say, ``synth'' or ``del''?}
\begin{equation}\label{quotientlenslaws}
\ell.\get \; (\ell.\lput \; t) \equiv_T t \text{, and } \ell.\lput \; (\ell.\get
\; s) \equiv_S s
\end{equation}
Quotient lenses are useful in situations where a programmer wishes for the
transformation defined by a lens to have the same behavior on data that differ
only in inessential details. For instance, a programmer may wish to ``quotient
away'' the number of white space characters between data items, the
capitalization of various strings, the sequence of fields in a record, or the
order of items in a list.

@@

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the edges,''
just like Boomerang's quotient lenses. 
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{QREs}
\caption{QRE lens from source $P$ to target $Q$.  $P$ and $Q$ are QREs, each
  consisting of a ``whole'' set ($W(P)$ and $W(Q)$) and a ``kernel'' part
  ($K(P)$ and $K(Q)$) Between the two kernels is a plain bijective lens.
  The \kw{get} function of the whole lens takes an argument from $W(P)$,
  applies the canonizer for $P$ to obtain a canonical representative in
  $K(P)$, then applies the \kw{get} of the plain lens, yielding an element
  of $K(Q)$, which is a subset of $W(Q)$.  The \kw{put} function does the
  reverse, mapping from $W(Q)$ to $K(P)$ (hence $W(P)$.  This lens is in
  normal form, with canonizers (specified by QREs) at the outer edges and an
  ordinary bijective lens in the center. \BCP{Both the figure and the
    caption are new---what do you think??}}\saz{I think that this figure is a
  big improvement.  One tiny possible tweak: color the \kw{get} and \kw{put}
  keywords in the same color as in the text.}\bcp{Can do, if we keep the
  color in the text :-).  How are we going to decide about that?}\saz{Vote?}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: Q_1 \Leftrightarrow Q_2$ where $Q_1$ and $Q_2$
are QREs. In the {\em get} direction, a QRE lens $q_1: Q_1 \Leftrightarrow Q_2$ uses the
source QRE $Q_1$ to compute a canonical representative for the data modulo the
equivalence relation defined by $Q_1$ and then applies the \kw{get} function of a
bijective string lens $\ell_1$ to this representative. In the {\em put} direction, $q_1$
operates similarly, but using the QRE $Q_2$ and the $\lput$ function of
$\ell_1$.

@@

\BCP{Wrote the rest---comments?}\saz{This seems OK---maybe swap the two
  paragraphs.  The second one sounds more incremental and less exciting than the
  first, so it's maybe not as good of a note to end on?}
Looking ahead, we would like to investigate whether the techniques proposed
here can be applied to synthesizing quotiented variants of richer classes of
lenses, such as ``classic'' asymmetric lenses~\cite{FosterLenses} or
symmetric lenses~\cite{symmetriclenses}.  The main technical issues are
developing synthesis procedures for the unquotientied variants of a given
class---we are currently attacking the case of symmetric lenses---and
showing that quotiented lenses in this class can be normalized along the
lines of Theorem~\ref{normal form}.
